---
title: "Pet Names in Seattle and NYC"
author: "Chelsea Li, Lomash Sharma, Samarth Kunhody"
date: "2023-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE,
                  	warning = FALSE)
```

```{r}
library(tidyverse)
library(dplyr)
library(lubridate)
library(stringr)
```

## Introduction:
Pet ownership is a fascinating lens through which we explore the cultural dynamics of Seattle.
In this project, we delve into the Seattle Pet Licenses dataset, sourced from the City of Seattle,
to unravel the nuances of pet ownership in the city. Our primary goal is to offer valuable insights for a diverse audience,
including pet enthusiasts, businesses, city planners, and researchers, by comparing and contrasting Seattle's data
with the New York City Dog Licensing dataset.  

Seattle, renowned for its diverse and pet-friendly community, stands as a unique backdrop for exploring the intricate relationship
between residents and their beloved pets. The city's commitment to fostering strong bonds between humans and animals creates a
captivating environment for investigation. The Seattle Pet Licenses dataset serves as our window into this world, offering a
comprehensive glimpse into the dynamics of pet ownership in the city.  

Existing Understanding:
As of October 12, 2023, our understanding of Seattle's pet landscape is shaped by the rich cultural significance attached to this
bond. The dataset encapsulates not only the variety of pets but also highlights the neighborhoods where these connections thrive.
However, beyond anecdotal evidence, there is a need to delve deeper into the data to uncover patterns, trends, and correlations that
might not be immediately apparent. From the media, common features of pets in peoples' daily lives shows that
they are important companions to people.

What We Are Doing:
Our project aims to unravel the intricate tapestry of Seattle and NYC's pet ownership dynamics by addressing some of these key
questions and storylines:
  
- What are the most popular pet names in Seattle and NYC?

- Are there names that are popular in one city, but less in the other?

- What is the breakdown of pet breeds in both cities?  
  
Why?
This is interesting because pets are a vital part of many modern citizens' lives, especially in big metro
cities like Seattle and NYC. Insights into the names and breeds of pets in these areas can be entertaining
to acknowledge. We and our audience can gain an understanding of the different pets involved in our and
our communities' lives. It can also give prospective pet owners information that may help them choose names
for their future pets.  

In essence, our project aims to transform raw data into a compelling narrative, shedding light on the diverse,
interconnected facets of pet ownership in Seattle and NYC. The findings are not just statistics; they represent a meaningful
exploration of the city's social fabric, providing valuable insights for various stakeholders,
from pet enthusiasts to city planners.  

## Data Explained  

```{r}
seattle <- read_delim("../final project/datasets/Seattle_Pet_Licenses.csv",
                      show_col_types = FALSE)
```

Original Seattle Dataset:
The [Seattle] (https://data.seattle.gov/Community/Seattle-Pet-Licenses/jguv-t9rb) dataset is a comprehensive compilation,
featuring key details such as license issue dates, unique license numbers, the names of pets, their species, and primary and
secondary breeds. Additionally, the dataset provides a geographic dimension with ZIP codes, allowing for a nuanced exploration of
pet ownership trends throughout the city. This data is Collected by Department of Finance and Administrative Services, The City of
Seattle, in order to provide demographic and information data to the general public. It is accessible through the city’s official
data portal, ensuring credibility and transparency. It’s available in the public domain and retrieved from Seattle Pet Licenses |
City of Seattle Open Data portal. Anyone is allowed to use this data without attribution for non-commerical uses. We believe this
data is suitable for our purpose because being an official document, it’s credible, and it includes a wide range of duration that
will help capture trends over time. In addition, its comprehensiveness in including details such as breed will help us illustrate
strong analytical points and patterns.  


Quality Issues: Seattle Dataset
According to the problem detecting function, there is an issue in
the 23773rd row and 7th column, where the expected variable is a double, but the actual
value is a character.  
```{r}
problems(seattle)
```


```{r}
seattle[23773, 7] <- 98104

head(seattle)
```

To fix this error, we opened the Excel CV file and found the exact value in that position,
and replaced it here in the dataset.  
```{r}
seattle[23773, 7] <- 98104

seattle[23773, 7]
```


Original NYC Dataset:
The [New York City Dog License] (https://data.cityofnewyork.us/Health/NYC-Dog-Licensing-Dataset/nu7n-tubp) dataset contains dog
names, gender, birth year, and breeds. The population is of licensed pets in NYC, and the sample is publicly licensed pets from 2014
onwards. It was compiled by NYC’s department of mental health and hygiene, for the purpose of keeping track of dog licenses. It
originated from NYC's official city website. This increases reliability because the information is coming from the original source.
The dataset is freely available on the internet and is allowed to be used non-commercially. We are using this dataset because it's
the official record from the city of NYC, and covers a widebreadth of detail and time.  

Quality Issues: NYC Dataset
According the the problems function, there are no value mismatching issues with this dataset.
```{r}
nyc <- read_delim("../final project/datasets/NYC_Dog_Licensing_Dataset_20231129.csv",
                  show_col_types = FALSE)
problems(nyc)
```


## Methods

We first standardized the capitalization for pet names for both datasets. This was done to ensure that
names will be accurately compared to each other later.  
However, it should be noted that a trade-off is that the originality of some names will be lost.
Perhaps there are names where the capitalization is a purposeful intent. But changing all
these names to the default title case will sacrifice some of that uniqueness. Still,
this is necessary to do in order to optimal comparison between the two datasets.

Standardizing capitalization for Seattle Dataset:

```{r}
seattle <- seattle %>% 
  mutate(`Animal's Name` = str_to_title(`Animal's Name`))

head(seattle)
```

Standardizing capitalization for NYC Dataset:
```{r}
nyc <- nyc %>% 
  mutate(AnimalName = str_to_title(AnimalName))

head(nyc)
```
  

We then standardized dates, turning character dates in DATE variables. While our final
analysis will not take TIME into account, we still want to clean up this variable
for the sake of organizing the individual datasets.

Standardizing dates for Seattle Dataset:
```{r}
seattle <- seattle %>% 
  mutate(date = str_replace_all(seattle$`License Issue Date`,
                                c("January" = "01", "February" = "02", "March" = "03",
                                  "April" = "04", "May" = "05", "June" = "06",
                                  "July" = "07", "August" = "08", "September" = "09",
                                  "October" = "10", "November" = "11", "December" = "12"))) %>% 
  mutate(Date = mdy(date))

head(seattle)
```


Standardizing Dates for NYC dataset:
```{r}
nyc <- nyc %>%
  mutate(Date = mdy(LicenseIssuedDate))

head(nyc)
```
  
The variables we are keeping: Date, Pet Breed, Animal Name.
We are keeping these variables because they directly pertain to our storyline questions.

Adjusted Column Names for Seattle:
```{r}
seattle <- seattle %>%
  mutate(ZipCode = `ZIP Code`) %>% 
  mutate(Breed = `Primary Breed`) %>% 
  mutate(Name = `Animal's Name`) %>% 
  select(Date, ZipCode, Breed, Name)

head(seattle)
```

Adjusted Column Names for NYC:
```{r}
nyc <- nyc %>%
  mutate(Breed = BreedName) %>%
  mutate(Name = AnimalName) %>% 
  select(Date, ZipCode, Breed, Name)

head(nyc)
```

The next adjustment is dataset size. We are cutting the rows down to 40,000 for both datasets.
This is to ensure that the data we use for analysis is the most recent data, and so that
the final dataset isn't too large.  
However, a trade-off of this is that not all the data will be analyzed. Some important trends will be lost.
But cutting down the size is necessary, otherwise the final dataset will be too overwhelming.  

Adjusting Size for Seattle:
```{r}
seattle <- seattle %>% 
  tail(40000)

head(seattle)
```

Adjusting Size for NYC:
```{r}
nyc <- nyc %>% 
  tail(40000)

head(nyc)
```
  
First Merging: Combining the cleaned individual datasets.  

```{r}
names <- bind_rows(seattle, nyc)
head(names)
```

New CATEGORICAL VARIABLE: City

```{r}
names <- names %>% 
  mutate(City = ifelse(substr(ZipCode, 1, 1) == 1, "New York City", "Seattle"))

head(names)
```


New Column: Number of Each Distinct Pet Name (In both NYC and Seattle)
```{r}
names <- names %>%
  count(Name) %>%
  mutate(Total = n) %>% 
  select(Name, Total)

tail(names)
```
  
New Dataset: Number of each distinct pet name in Seattle:
```{r}
seattleCount <- seattle %>% 
  count(Name) %>% 
  mutate(Seattle = n) %>% 
  select(Name, Seattle)
head(seattleCount)
```
  
New Dataset: Number of each distinct pet name in NYC:
```{r}
nycCount <- nyc %>% 
  count(Name) %>% 
  mutate(NYC = n) %>%
  select(Name, NYC)
head(nycCount)
```

  
Merging: Number of each name in total, in Seattle, and in NYC
```{r}
names <- left_join(names, seattleCount)

names <- left_join(names, nycCount)

names

names <- names %>% 
  mutate(Seattle = ifelse(is.na(Seattle), 0, Seattle))

names <- names %>% 
  mutate(NYC = ifelse(is.na(NYC), 0, NYC))

names <- names %>% 
  arrange(desc(Total)) %>% 
  filter(Name != "Unknown") %>% 
  filter(Name != "Name Not Provided") %>% 
  filter(Name != "None") %>% 
  filter(Name != "Name")

head(names)
```
  
New Categorical Variable: UniqueTo[City]
** Unique = at least 70% of the instances of a particular name comes from one city

```{r}
names <- names %>% 
  mutate(UniqueToSeattle = ifelse(Seattle / Total >= 0.7, TRUE, FALSE)) %>% 
  mutate(UniqueToNYC = ifelse(NYC / Total >= 0.7, TRUE, FALSE))

head(names)
```

  
New datasets: Numeric rankings of names in both cities  

Seattle:
```{r}
seattleRank <- seattleCount %>% 
  arrange(desc(Seattle)) %>% 
  mutate(Seattle = row_number())

head(seattleRank)
```
  
  
NYC:
```{r}
nycRank <- nycCount %>% 
  arrange(desc(NYC)) %>% 
  mutate(NYC = row_number()) %>% 
  filter(Name != "Unknown") %>% 
  filter(Name != "Name Not Provided") %>% 
  filter(Name != "Name")

head(nycRank)
```
  
New dataset: Overall numeric ranking of names (Seattle & NYC)
```{r}
namesRank <- names %>% 
  mutate(rank = row_number()) %>% 
  select(Name, rank)

head(namesRank)
```
  
Merging: Numeric ranking of names (Overall, Seattle, NYC)
```{r}
rank <- left_join(namesRank, seattleRank)

rank <- left_join(rank, nycRank)

rank <- rank %>% 
  mutate("OverallRank" = rank) %>% 
  select(Name, OverallRank, Seattle, NYC)

head(rank)
```
  
  
In order to analyze breeds, we are making new datasets focused on the variable BREED.
  
New dataset: Seattle Breeds
```{r}
seattleBreeds <- seattle %>% 
  count(Breed) %>% 
  arrange(desc(n))

head(seattleBreeds)
```
  
New Dataset: NYC Breeds
```{r}
nycBreeds <- nyc %>% 
  count(Breed) %>% 
  arrange(desc(n))

head(nycBreeds)
```


FINAL SUMMARIZATION TABLE:  

Each row represents a distinct pet name.  
Number of observations: 16,979  
Relevant variables 
- Name (chr): pet name  
- OverallRank (int): Rank of pet name, considering both Seattle and NYC  
- Seattle (int): Ranking of the name in Seattle  
- NYC (int): Ranking of the name in NYC  
- UniqueToSeattle (logical): Name is significantly more popular in Seattle  
- UniqueToNYC (logical): Name is significantly more popular in NYC  
  
Missing and non-plausible values:  
There were missing valuables for name such as "Name", "Name Not Provided", and "Unknown", for the NYC Dataset.
But since those were filtered out.  
Although, there are NA values for the names which don't exist in a certain city.
  
Total NA Values:
```{r}
rank %>% 
  filter(is.na(Seattle) | is.na(NYC)) %>% 
  nrow()
```
  
Total Seattle NA Values:
```{r}
rank %>% 
  filter(is.na(Seattle)) %>% 
  nrow()
```
  
Total NYC Values:
```{r}
rank %>% 
  filter(is.na(NYC)) %>% 
  nrow()
```
  
```{r}
unique <- names %>% 
  select(Name, UniqueToSeattle, UniqueToNYC)

rank <- left_join(rank, unique)

rank
```

  
## Results:

  
Seattle Breeds Chart:
This shows that the top 5 pet breeds in Seattle are Chihuahuas, Domestic Medium Hairs,
Domestic Shorthairs, Golden Retrievers, and Labrador Retrievers. Domestic Shorthairs make up
almost half, followed by Labrador Retrievers, followed by smaller portions of every other breed.  
```{r}
seattleBreeds %>% 
  head(5) %>%
  ggplot(aes(x = "", n, fill = Breed)) +
  geom_bar(stat = 'identity', width = 1, color="white") +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(title = "Top 5 Pet Breeds in Seattle")
```
  
  
NYC Breeds Chart:
This shows that the top pet breeds in NYC are Chihuahuas, Labrador Retrievers,
Labrador Retriever Crossbreeds, Shih Tzus, and Yorkshire Terriers.
Out of all these breeds, Yorkshire Terriers and Chihuahuas make up
the biggest portion, while the other 3 breeds equally make up the rest.  
```{r}
nycBreeds %>%
  filter(Breed != "Not Provided") %>% 
  head(5) %>%
  ggplot(aes(x = "", n, fill = Breed)) +
  geom_bar(stat = 'identity', width = 1, color="white") +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(title = "Top 5 Pet Breeds in NYC")
```

    
Top 10 Pet Names in Seattle:
The most popular pet names in Seattle are Bella, Charlie, Daisy, Lucy, Luna,
Max, Milo, Penny, Rosie, and Ruby. Visibly, Luna makes up the biggest proportion,
but the proportions of the other names are spread across considerably equally.
  
```{r}
seattleCount %>%
  arrange(desc(Seattle)) %>%
  head(10) %>%
  ggplot(aes(x = "", Seattle, fill = Name)) +
  geom_bar(stat = 'identity', width = 1, color="white") +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(title = "Top 10 Pet Names in Seattle")
```
  
Top 10 Pet Names in NYC:
The most popular pet names in NYC are Bella, Charlie, Coco, Daisy, Lola,
Luna, Max, Milo, Rocky, and Teddy. Luna, Max, and Bella make up slighty bigger portions
than the rest of the equally portioned names.
```{r}
nycCount %>%
  filter(Name != "Unknown") %>% 
  filter(Name != "Name Not Provided") %>% 
  arrange(desc(NYC)) %>%
  head(10) %>%
  ggplot(aes(x = "", NYC, fill = Name)) +
  geom_bar(stat = 'identity', width = 1, color="white") +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(title = "Top 10 Names in NYC")
```

  
City Specific Rankings of Top 5 Names:
This plot features the OVERALL top 5 pet names, and their respective numeric rankings
in Seattle and NYC. From this plot, we can tell that the relative rankings for
Bella, Charlie, Luna, and Max are not significantly different in Seattle compared to NYC.
However, Lucy is a mild outlier. While its ranking in Seattle is 3rd, its ranking in NYC 13.
In this context, that is quite a large difference.
```{r}
rank %>% 
  head(5) %>% 
  ggplot(aes(Seattle, NYC, col = factor(Name))) +
  geom_point(size = 3) +
  labs(x = "Seattle Ranking", y = "NYC Ranking", col = "Name", title = "Top 5 Pet Names Overall in Seattle and NYC")
```

  
Names from the Overall Top Ranking With Notable Popularity Differences
While the graph above begins to show that there are distinctions between individual
popularity rankings between a pet name in Seattle and NYC, it only features the Overall Top 5 Names.
To further analyze that, we made a bargraph to show the Overall Top 5 most popular names that
are unique to one city. In our case, we have defined "unique" to mean that at least 70% of the total count
of a certain name comes from either Seattle or NYC. As displayed below, the Overall Top 5 names with
big gaps in popularity between these two cities are: Coco, Lucky, Mia, Rocky, and Teddy.
Interestingly, for all of these names, they are unique to NYC. From this, we can deduce that NYC has a large
effect on the rankings for the overall top names. This applies especially for the name Mia, which ranks at ~80 for Seattle,
but at ~20 for NYC. NYC's high ranking for this name boosts its overall ranking despite Seattle having a low ranking for it.
```{r}
rank %>% 
  pivot_longer(c(Seattle, NYC), names_to = "City", values_to = "Rank") %>% 
  select(Name, City, Rank, UniqueToSeattle, UniqueToNYC) %>% 
  filter(UniqueToSeattle == TRUE | UniqueToNYC == TRUE) %>%
  head(10) %>% 
  ggplot(aes(x= Name, y = Rank, fill = factor(City))) +
  geom_bar(stat = 'identity', position = 'dodge') +
  labs(x = "Name", y = "Rank", fill = "City", title = "Names with Notable Popularity Differences")
```
    
Biggest Popularity Differences in the Top 1000 Overall:
We want to dive deeper into regional differences for popular pet names. For this chart, we are
taking the Overall Top 1000 pet names in Seattle and NYC, and finding the 5 names within that,
with the biggest difference in popularity. From below, we can see that these 5 names are:
Cedar, Lewis, Minerva, Neko, and Rainier. Intriguingly opposite the last graph, the 5 names here
are significantly more popular in Seattle. The difference between city rankings for each name
ranges from ~4000 to ~7700. These are large numbers. It makes sense that these 5 names are much
more popular in Seattle than NYC because the names are mostly nature themed and are words
associated with the Pacific Northwest. This is especially the case for Cedar, which is a common tree
here in the Seattle, and Rainier, which most likely is related to Mount Rainier here in Washington.
This is important because if these 5 names were not ranked so high in Seattle, their overall rankings
would not have made it to the Top 1000.  
```{r}
rank %>%
  head(1000) %>% 
  select(Name, Seattle, NYC) %>% 
  mutate('Difference(SEA-NYC)' = Seattle - NYC) %>%
  arrange(`Difference(SEA-NYC)`) %>% 
  head(5) %>% 
  ggplot(aes(x = Seattle, NYC, col = factor(Name))) +
  geom_point(size = 3) +
  labs(x = "Seattle", y = "NYC", col = "Name",
       title = "Names from the Overall Top 1000 With Significant Popularity Differences")
```
   
## Discussion  

Our guiding questions were:  

- What are the most popular pet names in Seattle and NYC?  

Our findings tell us that the most popular pet names in Seattle and NYC,
line up fairly similar with the national rankings. All of the names in the Top 10
for both cities, include some of the most popular pet names across the nation.
Most notably, Luna, Bella, Max, and Charlie. These pet names have been around
for a long time, which tells us that pet owners may lean toward more classic, timeless, names.
These claims can generally extended for the entire pet population of Seattle, because
in personal experience as well as anecdotes online, these highly popular pet names
commonly show up over and over again. However, it's important to note that these pet names
are common *American* pet names. Families from different countries may be naming their pets with names
from their culture, but this data does not reflect that. Different neighborhoods will represent different
ethnic demographics who may choose to name their pets differently. In order to further analyze this,
we would need to investigate into pet names by neighborhood.  


- Are there names that are very popular in one city, but much less in the other? 
  
The data tells us that while the overall rankings can tell us general information
about which pet names are the most popular, there are regional differences in
name choices that should be considered. For example, Seattle pet owners are much more
likely to choose names that are related to wildlife and nature, than NYC pet owners are.
We believe this is generally applicable to most pet owners in their respective cities.
It makes sense that names that are associated with specific aspects of one area will
be used more often in that area, than other areas. To analyze this further, we would need to
study the latter end of the dataset in greater detail, where more regional specificpet names are placed.
  
- What is the breakdown of pet breeds in each individual city?
  
The most popular pets in Seattle are cats and dogs, while the most popular
pets in NYC are dogs. Domestic short hair cats are very popular in Seattle, and so are
variations of retriever dogs, which are popular in New York City too. This generally can be applied
for pet owners in Seattle, NYC, and every city. Dogs and cats are very common pets in households.
However, it is mildly perplexing that no singular cat breed shows up in the Top 5 Breeds for New York City.
While we may tentatively deduce that dogs may be more popular in NYC, we don't firmly believe this dataset reflects
the make-up up of pet species in New York. Perhaps for some reason, pet owners are more likely
to license their dogs than they will with their cats. To understand this further, we would need
a more comprehensive data set that potentially encompasses even more pets in New York City.
  
## Summary  

Pets are very valued parts of the lives of city goers, particularly in Seattle and New York.
This is shown by how the sheer amount of pets that are constantly being adopted into new homes,
depicted by the wide expanse of these two datasets from the cities of Seattle and New York City.
The diverse selection of names that families choose for their pets shows how these pets individually
have unique personalities and dynamics within their home. While some of the most classic pet names
remain the most popular to this day such as Bella and Max, there are a huge variety of distinct names
that some pets are given. There are thousands of unique names per city, and each of them tells a story
about them as a pet, and who they are to their families. Some names even reflect the city that they're from,
a special tribute to the culture they are brought up in. This is why the popularities for certain names
are much more popular in Seattle than in New York, or vice versa. Just like languages and human names,
they reflect specific cultures, times, and heritages, in different communities. Ever since dogs and cats
were domesticated to become human companions, they have been a treasured part of humans' lives, and through
the display of cats' and dogs' popularity in these data, we can look forward to them continuing to
be an uplifting and honored part of society.

Understanding popular pet breeds in urban areas aids designers and planners in creating tailored, pet-friendly spaces. This knowledge ensures the development of recreational areas suitable for specific breeds, adequate exercise facilities, and pet-friendly infrastructure like walking paths and waste disposal stations. It also allows for the design of spaces promoting positive socialization, implementing safety measures, and ensuring inclusivity for pets of various sizes and temperaments. By considering the preferences of popular breeds, urban spaces become more engaging, inclusive, and enjoyable for both pets and their owners.
